---
title: "Compute Moran's Index"
author: Lauren Hsu^[Harvard School of Public Health]
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Compute Moran's Index}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

See the source code of this vignette for the script used to compute the Moran's I values analyzed in [2. Tumor spatial autocorrelation and clinical prognosis](https://laurenhsu1.github.io/BIRSBIO2020.scProteomics.exploratory/articles/moran_clinical.html).

```{r packages, eval = F, include = F}
library(sf)
library(stars)
library(raster)
library(adespatial)
library(maptools)
library(spdep)
library(stringr)
library(readr)
library(dplyr)
```

## Image loading and spatial processing

Download the mibiTOF data from the [Angelo Lab website](https://www.angelolab.com/mibi-data). 
Move the datasets into the `data` directory.

This portion of the script converts each image into a more analysis-friendly format, and computes Gabriel graphs/neighborhoods, which is used as the spatial weighting matrix (SWM) in the Moran's I calculation below.

*Some parts of this image processing steps of this script are adapted from code by Pratheepa Jeganathan and Kris Sankaran. Github: @PratheepaJ @krisrs1128*

```{r, eval = F, include = F}
DATAPATH <- "../data"
SAVEPATH <- "../data/spatial_proc"

data_dir <- file.path(DATAPATH, "TNBC_shareCellData")

process_tiffs <- function(dirname){
  fns <- list.files(data_dir, str_interp(".tiff"))
  moran_list <- list()
  
  for(fn in fns){
    samp_list <- list()
    
    f <- paste(data_dir, fn, sep = '/')
    
    samp_id <- str_split_fixed(fn,'_',2)[1] # with p
    sample_id <- as.numeric(str_split_fixed(str_split_fixed(fn,'_',2)[1],'p',2)[2]) # without p as numeric
    samp_list[['samp']] <- samp_id
    
    polys <- read_stars(f) %>%
      st_as_sf(merge = TRUE) %>%
      st_cast("POLYGON")
    colnames(polys)[1] <- colnames(cell_data)[2]
    
    cell_data <- read_csv(file.path(data_dir, "cellData.csv")) %>%
      filter(SampleID == sample_id)
    
    polys <- polys %>%
      inner_join(cell_data) %>%
      group_by(cellLabelInImage) %>% # some regions get split into two adjacent polys --> merge
      summarise_all(first)
    
    samp_list[['polys']] <- polys
    
    nb.maf <- poly2nb(polys)
    xy <- polys %>% st_cast("POINT") %>% st_coordinates()
    nbgab <- graph2nb(gabrielneigh(xy), sym = TRUE)
    
    samp_list[['nb.maf']] <- nb.maf
    samp_list[['xy']] <- xy
    samp_list[['nbgab']] <- nbgab
    
    nblgab <- nb2listw(nbgab)
    distgab <- nbdists(nbgab, xy)
    
    samp_list[['nblgab']] <- nblgab
    samp_list[['distgab']] <- distgab
    
    saveRDS(samp_list, file = paste0(SAVEPATH,'/', samp_id,'.rds'))
  }
}

process_tiffs(dirname = data_dir)
```


## Computing Moran's Index

This portion reads in the processed files generated above and computes Moran's I for each variable, which includes attributes like protein expression values (Moran's I is computed for each), cell type dummy variable, and cell size.

The output of this code is the `resMoranTest.csv` file included in the `data` folder.

```{r, eval = F, include = F}
rdsfolder <- SAVEPATH
pFiles <- file.path(rdsfolder, dir(rdsfolder))

runMoranScript <- function(pFiles){
  pFilesNames <-  reshape2::colsplit(reshape2::colsplit(pFiles,"/",1:4)[,4],"\\.",1:2)[,1]
  
  res<-lapply(pFiles,sc)
  cols <- unique(unlist(sapply(res, names)))
  
  resM <- matrix(NA, ncol=length(cols), nrow=length(pFilesNames), dimnames = list(pFilesNames,cols))
  for (i in seq_along(res)) resM[i,names(res[[i]])]<-res[[i]]
  return(resM)
}

sc<-function(pFile) {
  print(pFile)
  p<-readRDS(pFile)
  if (!exists("p")) Sys.sleep(5)
  print(ls())
  mp<- moranS(sf=p)
  rm(p)
  return(mp)
}

moranS<-function(sf){
  # Exclude non number cols or where max-min =0
  to_drop <- c("cellLabelInImage" ,"SampleID" , "geometry", # non numeric
               'Na','Si','P','Ca','Fe','Background', 'dsDNA','Ta','Au',  # elemental / irrelevant
               'tumorCluster', 'immuneCluster','immuneGroup','Group') # factors
  p <- sf$polys
  nbl <- sf$nblgab
  n <- colnames(p) [! colnames(p) %in% to_drop]
  n <- n[!apply(p[, n,drop=TRUE],2, function(x)max(x)-min(x))==0]
  tt <- lapply(n, function(x) (moran.test(p[,x, drop=TRUE],nbl)))
  ms <- sapply(tt, function(x) x$estimate[1])
  names(ms) <- n
  return(ms)
}

resMoranTest <- runMoranScript(pFiles)
write.csv(resMoranTest, file=file.path(SAVEPATH,"resMoranTest.csv"))


```
